---
title: "filmPredictionML"
author: "Jared Bowden"
date: "10/20/2014"
output: html_document
---

Data source: http://grouplens.org/datasets/movielens/

## Preprocessing
Nuke it from space 
```{r remove objects}
# Remove all objects from memory
rm(list=ls())
```

Read in the data 
```{r Read in data}
# Set the working directory
setwd("~/Dropbox/dataIncubator/project//hetrec2011-movielens-2k-v2/")

# Read in the file as a csv
firstData <- read.table("./user_ratedmovies-timestamps.dat", 
  header = TRUE, colClasses = "character") 

# THE WAY BACK
movieData <- firstData
```

These time stamps are kinda confusing...

Alright, we have a bunch of 13 digit numbers that are not converting to dates.
Oh. I think this is UTC. We can deal with this.
```{r debug the time stamps}
# Some indication that we might need to divide by 1000. Let's run a test
# case: 1223625490000
format(as.POSIXct((1223625490000/1000), origin='1970-01-01'), 
       format='%Y/%m/%d %h')

# Looks good. Let's convert the rest of the dates.
movieData$dateInfo <- as.numeric(movieData$timestamp)/1000

movieData$date <- format(as.POSIXct(movieData$dateInfo,origin='1970-01-01'), 
       format='%Y/%m/%d')

# Confirm that this is looking good
head(movieData)
tail(movieData) # Looks good

# Convert these values to dates
movieData$date <- as.Date(movieData$date)
```

Let's review the integrity of the data set.

Quick review of the data
```{r Look at the column names}
# Check the column names 
names(movieData)

# and the data structure. 
str(movieData) # There are a few data types in here that will need to change

# Convert userID, movieID, and rating values to numeric data type.
movieData$userID <- as.numeric(movieData$userID)
movieData$movieID <- as.numeric(movieData$movieID)
movieData$rating <- as.numeric(movieData$rating)

# Confirm that data type changes have been applied
str(movieData)
```

Check for NA values
```{r NA values}
# Check for NA values. 
colSums(is.na(movieData)) # Looks good
```

Now, I'll want to parse a few more data points from this date information.
```{r parse date data}
library(lubridate)

# Hmm... It looks like all of this date information is out of order. Let's
# fix that
movieOrder <- movieData[order(movieData$date, decreasing = FALSE), ] 

# Let's start by parsing the month that the review was made.
movieOrder$dateMonth <- month(movieOrder$date, label = T)

# I'm also interested in adding week-of-the-year information, as well as year.
# We'll need this to forecast a time series, later.
movieOrder$dateWeek <- week(movieOrder$date)
movieOrder$dateYear <- year(movieOrder$date)

# Confirm that this is looking good
head(movieOrder)
tail(movieOrder) # Looks good

# Date range for the dataset appears to be 1997/09/17 - 2009/01/04
```

## Exploratory data analysis

The first question I want to address: do average movie ratings change across
months? To put this another way: is there any truth to the notion that summer
films poorly rated, and increase steadily leading up to Oscar-season?
```{r ratings by month}
# Let's calculate average rating by month
movieMonth <- aggregate(rating ~ dateMonth, data = movieOrder, mean)

# Let's graph this information
library(ggplot2)

ggplot(movieMonth, aes(group = 1, x = dateMonth, y = rating, 
    color = movieMonth$dateMonth, fill = movieMonth$dateMonth)) +
    geom_point(size = 7) +    
    geom_line(color = "grey") +
    ylab("Average rating (out of 5)") +
    xlab("Month") +
    ggtitle("Movie ratings from 1997 - 2008, averaged by month") +
    theme(axis.text.x = element_text(angle = 270, hjust = 1, size = 12), 
          legend.title = element_blank(),
          axis.text.y = element_text(size = 12)) 
```

The second question I want to address: can we use seasonal patterns in movie
ratings to accuratly forcast future ratings?
```{r time series forecast}
library(forecast)

# Let's calculate average rating by month and year.
movieMY <- aggregate(rating ~ dateYear + dateMonth, data = movieOrder, mean)

# Now, sort the resulting data frame by time.
movieTS <- movieMY[order(movieMY$dateYear, 
                            movieMY$dateYear, decreasing = FALSE), ] 

# But to do this, we are going to need complete years. This means dropping 
# January of 2009, and Sept - Dec. 1997
subTS <- movieTS[5:(length(movieTS$dateYear) - 1), ]

# Now that this is is sorted, convert to time, and paste these values. 
subTS$dateMonth <- month(as.numeric(subTS$dateMonth), label = T)
subTS$date <- paste(subTS$dateMonth, subTS$dateYear)

# Make a time series
timeSeries <- ts(subTS$rating, frequency = 12, start = c(1998, 1))

# Forcast from 2008 - 2015
HWtimeSeries <- HoltWinters(timeSeries)
HWtimeSeries2 <- forecast.HoltWinters(HWtimeSeries, h = 72)

# And plot this forecast
plot.forecast(HWtimeSeries2, ylab = "Average rating (out of 5)",
              xlab = "Date (years)",
              main = "Forecasted movie ratings from 2008 to 2015")
```

## Machine learning
Future directions.




